A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-03-14 paper/bare_conf.tex
Minor fixes

% Register allocation and register locality
\subsection{Domain-drive Optimization of Register Management}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). 
In this context, register tiling is a critical optimization. A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP. This is automated by some vendor compilers~\cite{comdex}. Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected. Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture. Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning is not an attractive approach since it would be performed at run-time when kernel code is available. The optimization strategy is described in the following, along with an overview of our compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, the $split$ and $op$-$vect$ transformations may be introduced; their implementation is based on analysis and trasformation of the AST. The compiler handles any possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used (if a factor is 0, then the corresponding transformation is not applied). The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of ``free'' registers, i.e. those not used for keeping hoisted terms (line 9-11).

Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance. Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.  We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors. This choice also simplifies the compiler's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying unroll-based optimizations, even if specific pragmas are added.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to accumulating increments in a register along the $i$ loop. The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.

>>>>

% Register allocation and register locality
\subsection{Domain-driven Vector-register Tiling}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In the compiler we have developed, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers can be introduced as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a domain-specific vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected. Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture. Because of the large number of (possibly parametric) transformations, we need a mechanism to select the most suitable optimization strategy for a given problem. Autotuning might be used, although at the moment we avoid it to minimize the run-time overhead. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of our compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, the $split$ and $op$-$vect$ transformations may be introduced; their implementation is based on analysis and trasformation of the AST. When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRbi) intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available vector registers ($n\_regs$) and the number of variables iterating along the $j$ and $k$ dimensions ($n\_consts$ for independent variables, $n\_outer\_arrays$ for $j$ variables, and $n\_inner\_arrays$ for $k$ variables, assuming the $ijk$ loop order) to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used (if a factor is 0, then the corresponding transformation is not applied). The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. Finally, the unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e. those not used for storing hoisted terms (line 9-11).

Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters, like the unroll factor, becomes of great importance. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.  We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors. This choice also simplifies the compiler's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying unroll-based optimizations, even if specific pragmas are added.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations $ijk$ and $ikj$ are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between $ijk$ and $ikj$ depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are availabe to carry out the computation of the stiffness matrix.
